---
title: "Maximizing Elevation in the Himalayas"
author: "Carter Burr, Jack Schreuder, Alex Friedberg"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE, echo = FALSE}
library(tidyverse)
library(openintro)
library(broom)
library(lubridate)
library(tidymodels)
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv')
```

## Introduction


SHOULD MAKE AN INTRODUCTION MORE ABOUT THE TOPIC THAN ABOUT THE DATASET, Talk about how we are deciding to just use Everest to keep consistent data, and how it still has enough datapoints. 

For our final project, we decided to use a dataset compiled by Alex Cookson, where he first posted it to his blog at https://www.alexcookson.com/post/analyzing-himalayan-peaks-first-ascents/ , which we were able to then later find in the tidytuesday
github repository at https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md. 

This data was collected by [The Himalayan Database](https://www.himalayandatabase.com/), which is a complication of over 10,000 climbs of over 450 peaks in the Nepal Himalayas starting from 1905 to 2019. The individuals of the data are each different climber that attempted a climb between 1905 and 2019, including the hired helpers that would assist on each climb. When looking at the different variables with this data set, we can see that each expedition and mountain is identified, as well as the year and season with which the climb took place. This data also includes the dates that the expedition left base camp, as well as when they reached their high point and their entire journey was over.  



### Scientific Context

For mountain climbing enthusiasts, there is no greater story to tell than the one of how high they climbed in the Himalayan mountains. But climbing these uniquely enormous mountains is more than just pride, about one quarter of peaks in the Himalayan mountain range have never been climbed before. The human sense of adventure is insatiable, and not even the dangers of altitude sickness, extreme temperatures, and ever changing terrain can dissuade people from trying to climb up these mountains. Given the high level of danger and difficulty involved, it is important for climbers to understand factors that contribute to a successful climb, because if they don't the trip could be deadly for them and their sherpas. So how do climbers achieve the highest height they can? That is what we aim to answer with our analysis of this data.


### Codebook

Name of Variable   | Explanation
-------------------|-------------
expedition_id      | A unique chr that corresponds to each expedition
peak_name          | The name (chr) of the peak that this expedition is climbing
year               | The year (dbl) during which the expedition happened
season             | The season the expedition happend in as a chr, either 'Spring', 'Autumn', 'Winter', or 'Summer'
highpoint_metres   | A dbl expressing the highest point (in meters) the expedition reached before termination
members            | A dbl showing the number of non-hired members in the expedition
member_death       | The number (dbl) of members who died during the expedition
hired_staff        | The number (dbl) of hired staff, aka sherpas, who went on the expedition
hired_staff_deaths | A dbl expressing the number of hired staff who died during the expedition
oxygen_used        | A lgl variable where True means at least one member or hired staff used oxygen at any point, and False means that nobody ever used oxygen
trekking_agency    | The chr name of trekking agency from which the hired staff where hired. Appears as NA if no staff were hired


### Glimpse of our data
```{r codeblock1, echo = FALSE}
glimpse(expeditions)
```


```{r refiningData, echo = FALSE, warning= FALSE}
expeditions <- filter(expeditions, peak_name == "Everest")
```


## Model Building

Need 3 different models that measure 3 different explanatory variables for our response variable each. Need to brainstorm which models would be the best. 


```{r splittingData}
set.seed(3521)

expeditions_split <- initial_split(expeditions, prop = 0.50)

expeditions_train <- training(expeditions_split)
expeditions_test <- testing(expeditions_split)
```
*This can be what we use for our 50-50 data split, we can change the seed as long as we keep it consistent. 


### Model 1
#### Model Proposed by Carter
```{r model1_explanation, warning=FALSE, echo = FALSE}

graph1 <- expeditions_train %>% ggplot(mapping = aes(x = members,y = highpoint_metres)) +
  geom_point() +
  geom_smooth(method = "loess", formula = y ~ x)

graph2 <- expeditions_train %>% ggplot(mapping = aes(x = oxygen_used, y = highpoint_metres)) +
  geom_boxplot(outlier.colour="black", outlier.shape=4,
             outlier.size=2, notch=FALSE)

graph3 <- expeditions_train %>% ggplot(mapping = aes(x = season, y = highpoint_metres)) +
  geom_boxplot(outlier.colour="black", outlier.shape=4,
             outlier.size=2, notch=FALSE)



graph1+graph2+graph3
  


```

#### Interpretation
When deciding on what 3 explanatory variables I would be using for my model in order to best predict my response variable, I first started out with some background research to get a few ideas. First off, I looked at oxygen_used, and found many benefits to including the use of oxygen while reaching very high elevations. I also found research on the differences of climbing season. These looked like two very good categorical variables, I needed a numeric. To further look into this, I decided to make 3 seperate graphs with each of the variables I wished to use and saw how they individually looked against highpoint_metres. First off, I started with a point graph and looked at the correlation between members and highpoint, and saw that as the members increased the highpoint would also increase, but not by very much amount. Next, I looked at boxplots between if oxygen was used and its correlation to highpoint. Looking at this plot, we can clearly see that we would expect those who use oxygen to reach a higher point, as they appear to have an average and inter-quartile range much higher than those who do not use oxygen. Last off, I look at the effect season had on the high point, and saw some interesting results. First off, it appears we would get the highest on average with the least amount of range in Autumn. In spring the median is the same as autumn, but has more range in the 25%-50% quartile. Winter shows a slightly lower median, and then lastly we see summer with the lowest. These charts help us better understand what values and correlations to expect when we are making our model.  



```{r model1, echo = FALSE, warning = FALSE}

expeditions_model <- lm(highpoint_metres ~ members + oxygen_used + 
                        season, data = expeditions_train)
tidy(expeditions_model)

expeditions_pred <- predict(expeditions_model, newdata = expeditions_test) %>%
  bind_cols(expeditions_test %>% select(highpoint_metres)) %>%
  rename(pred = ...1)

expeditions_pred

glance(expeditions_model) %>% select(r.squared, adj.r.squared)


rmse(expeditions_pred, truth = highpoint_metres, estimate = pred)
```

#### Explanation:

##### Main Model Equation
Highpoint_metres = 7667.79 + 22.53*Memebers + 285.62*Oxygen_usedTRUE + 5.47*WinterYes - 58.25*SpringYes - + 245.23*SummerYes

##### Main Model Interpretation:
When looking at the main model equation that we made, we can intrepet a few things. First off, our intercept is 7667.79, meainging if you put in 0 for all of the values, we would expect the climber to make it up to 7667.79 meters. Now when we look at memebers, we can see a coefficient of 22.53, meaning that with each member you add, you are expected to reach 22.53 meters higher on average. Although this does not seem very significant, some parties can have 30 to 40 members, which gives a huge boost to the height you would be expected to reach. Looking at the oxygen categorical variable, we can see that if you use oxygen you are expected to reach 285.62 meters higher on average, which is a pretty significant value. Looking at the seasons, we can see that it is using autumn as a baseline, so for that season you do not add or subtract anything. For the summer on average you would lose 245.23 meters on your high point, in the spring you would lose 58.25 meters on average for your high point, and finally in the winter you would expect to gain only 5 meters on average to your high point. 

##### R^2 and RSME Values
Looking at our model, we get a R^2 value of 0.123, and an adjusted R^2 value of 0.117. These values are lower then what we would like to see, and it means that our indpendent variables have less correlation with the response variable than we would like. When we look at our RSME we get a value of 545.19. This value is a lot higher than we would like to see, as it shows us that we are experiencing a lot of error in our prediction model. Although this value is very high, when we look at the dataset, highpoint_metres is measured in thousands of meters, which is something we have to take into account when viewing this value. Although it is still higher than what we would like to see, it is not tremendously horrible when looking at the scale of the dataset. 


### Model 2
<<<<<<< HEAD
```{r model3, echo = FALSE, warning = FALSE}
Jack
```

Explanation


### Model 3
```{r model3, echo = FALSE, warning = FALSE}
Alex
```

Explanation
